### 2024-06-05 paper.tex
editorial main text changes

Here, we demonstrate that this coarse-graining approach can lead to signal dilution.
As an alternative, we propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
Our approach enables better discrimination of cell types and other features of interest and reduces the need for large numbers of cells.
We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.

To facilitate the analysis of scBS data, we have developed a software tool called \textit{MethSCAn} that implements these methods and provides additional functionality.

@@

Here, we discuss bisulfite sequencing, a method to study DNA methylation.
In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
After sequencing, these conversions allow for the determination of the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
Bisulfite sequencing can also be performed at single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{scMTseq,Clark2018}.

In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
We suggest improvements to current approaches, and demonstrate their value in benchmarks, using four real-world datasets.
Furthermore, we discuss how to perform comparative analyses.
Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis, using the described methods.

\subsection*{Standard approach}

@@

We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data. 
Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
A tutorial that showcases the analysis of a small example data set can be found at \url{https://github.com/anders-biostat/MethSCAn}.

@@

We also presented and published an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep{argelaguet2020mofa} (included in our benchmarks) or LIGER \citep{welch2019single}.
Once a low dimensional embedding is obtained, one can switch to well established methods from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc., e.g., by using either overall toolkits like Seurat \citep{seurat5}, ScanPy/EpiScanPy \citep{Wolf_2018,danese2021episcanpy} or any of the many available tools for specific tasks.
By offering these functions, \textit{MethSCAn} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Extended Data Fig.~\ref{fig:umaps}B).

>>>>

Here, we demonstrate that this coarse-graining approach can lead to signal dilution.
We propose improved strategies to identify more informative regions for methylation quantification, and a more accurate quantitation method than simple averaging.
Our approach enables better discrimination of cell types and other features of interest and reduces the need for large numbers of cells.
We also present an approach to detect differentially methylated regions (DMRs) between groups of cells, and demonstrate its ability to identify biologically meaningful regions that are associated with genes involved in the core functions of specific cell types.

Finally, we present the software tool \textit{MethSCAn} for scBS data analysis: \url{https://anders-biostat.github.io/MethSCAn}

@@

Briefly, in a bisulfite sequencing assay DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion.
After sequencing, these conversions allow for the determination of the methylation status of all cytosines covered by reads \citep{Frommer_1992}.
Bisulfite sequencing can also be performed at single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{scMTseq,Clark2018}.

In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data.
We suggest improvements to current approaches, and demonstrate their value in benchmarks, using four real-world datasets.
Furthermore, we discuss how to perform comparative analyses.
Finally, we present \textit{MethSCAn}, a comprehensive software toolkit to perform scBS data analysis.

@@

We have implemented the methods just described in a Python package with a command line interface, \textit{MethSCAn} (Single-Cell ANalysis of METHylation data), which also offers a number of other functionalities for the analysis of scBS data. 
Fig.~\ref{fig:workflow} illustrates a typical scBS data analysis with \texttt{MethSCAn} and provides an overview of the implemented core functionalities.
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or BISCUIT \citep{biscuit}, which are first stored in a compressed format that enables efficient access to all CpG sites of the genome (see Methods).
A tutorial that showcases the analysis of a small example data set can be found at \url{https://anders-biostat.github.io/MethSCAn}.

@@

We also presented an open source software tool, called \textit{MethSCAn}, that provides an easy-to-use implementation of the described algorithms.
It can start directly from the output of methylation callers such as Bismark \citep{bismark}, Biscuit \citep{biscuit} and methylpy \citep{methylpy} and produce a cell\texttimes region matrix.
We suggest iterative PCA as an approach to map the count matrix to a reduced-dimensional space overcoming the abundance of missing values.
Alternatively, one may use for this purpose established tools based on matrix factorisation such as MOFA+, \citep{argelaguet2020mofa} (included in our benchmarks) or LIGER \citep{welch2019single}.
Once a low dimensional embedding is obtained, one can switch to well established methods from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc., e.g., by using either overall toolkits like Seurat \citep{seurat5}, ScanPy/EpiScanPy \citep{Wolf_2018,danese2021episcanpy} or any of the many available tools for specific tasks.
By offering these functions, \textit{MethSCAn} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.
Our implementation can handle datasets of various sizes up to a hundred thousand cells (see Extended Data Fig.~\ref{fig:umaps}B).
